Modularly Composing Typed Language Fragments

### 2014-11-04 att-pldi15/att-pldi15.tex
Updated TR, related work, cleaned up obdy

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}\vspace{-3px}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}\vspace{-8px}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}. This form is analagous to the introductory form for kind $\kity$ described in Sec. \ref{sec:type-translations}, $\sqity{\qity}$. Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and dynamics simply recurse through these in the same manner as shown in Sec. \ref{sec:type-translations}. There is also an analagous unquote form, $\quq{\st}$. %The supplement gives the analagous rules.
The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$. These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.  Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$. We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
\[\small
\begin{array}{l}
\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}\\
\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
\quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx}\\
\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}\\
\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}\\
\quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
}\end{array}
\]

@@

The \emph{argument interface list} is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is the argument interface $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$. It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$.

@@

\inferrule[k-syn]{
    n' < n
}{
    \sofkX{\ssyn{n'}}{\kty\times\kitm}
}
\end{mathpar}
The following lemma characterizes thus characterizes the argument list interface generation judgement:
\begin{lemma}
\small If $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
\end{lemma}

The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$. This is to prevent out-of-bounds errors: tycon providers do not write these forms directly, only accessing them via the argument interface list, which is guaranteed to have the correct length. This is  safe because $n$ is only an upper bound:
\begin{lemma}
\small If $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and $n > n'
$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
\end{lemma}

@@

The selectively abstracted term translation judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement. The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous; see supplement):
\begin{mathpar}\small
\inferrule[abs-anatrans-new]{
    n \notin \text{dom}(\memG)\\
    \keyw{nth}[n](\es) = e\\
    \eanaX{e}{\st}{\iota}\\
    \tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\ity}{\memD'}\\
    (x~\text{fresh})
}{
    \edeabs{c}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
}
\end{mathpar}

@@

Finally, the twelfth premise checks the abstracted term translation against the abstracted type translation. Here, we are checking $\Delta_\text{ex/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.: 
\[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\]
In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with all references to term and type translations of regular strings replaced by variables and type variables, respectively. Because our intro opcon definition treated arguments parametrically, the check succeeds. We  describe  an ill-behaved operator in Sec. \ref{metatheory}.

@@

The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstracted. term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.

@@

\begin{theorem}[Static Preservation]\label{thm:static-preservation}
If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
\end{theorem}
The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be mutually defined. The mutual induction is well-founded because the total number of argument lists in the terms being considered decreases.
\begin{theorem}[Type Synthesis]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
\end{theorem}
\paragraph{Type Safety}
Type safety in a typed translation semantics requires that well-typed external terms translate to well-typed internal terms. Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong. To prove this, we must actually prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}:% produced as the translation of the external type. Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.

\begin{theorem}[Type Preserving Translation]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
\end{theorem}
\begin{proof-sketch}
We induct on the typing judgement. The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption. The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\emptyset \vdash \delta : \Delta_\text{abs}$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e. that all variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and so applying them in the conclusions of the rules gives a well-typed term.
\end{proof-sketch}

@@

\begin{tyconinvariant}[Regular String Soundness]
If $\eana{\emptyset}{\Phi_\text{rstr}}{e}{\sty{\tcvar{rstr}}{\concrx{r}}}{\iota}$ and $\iota \Downarrow \iota'$ then $\iota'=\texttt{"s"}$ and $\texttt{"s"}$ is in the regular language $\mathcal{L}(\texttt{r})$.
\end{tyconinvariant}
\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can simply treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$. Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}. The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type. But if so, the abstracted translation must have type $\tau_\text{abs}=\alpha$. Thus, invariants cannot be violated by relational parametricity in the IL (i.e. a ``free theorem'') \cite{WadlerThms}. \end{proof-sketch}

The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant, i.e. \emph{translation independence} of tycons, turns out to be the same reason extending the tycon context conserves all tycon invariants. A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g. $\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g. $\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, which would fail. %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
 We can state this more generally:

\begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
\item If $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\sttyidx, \iota)$.
\end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\sttyidx, \iota)$.
\end{itemize}
(if proposition $P(\sttyidx, \iota)$ is \emph{modular}, defined below)
\end{theorem}
\begin{proof-sketch}
The proof maps every well-typed term under $\Phi'$ to a well-typed term under $\Phi$ with the same translation, and if the term has a type constructed by a tycon in $\Phi$, e.g. $\tc$, the mapping still has a type constructed by that tycon with the same type translation. Thus we can apply the original tycon invariant. The only condition on $P(\sttyidx, \iota)$ is that it be preserved under this mapping on $\sttyidx$. We call such propositions \emph{modular}. Non-modular propositions are  uninteresting because the mapping's only effect on static terms is to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for $m$ uniquely associated with $\tc'$. On external terms, the mapping replaces all intro and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices. Typing, kinding, static normalization and the selective translation abstraction judgements are all preserved under this mapping, defined inductively in the supplement. \end{proof-sketch}
%Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$. Had we not gone through the machinations of holding types and terms abstract, however, the check would succeed, because $\keyw{str}$ would no longer have been held abstract as $\alpha_1$. This form of ``type translation independence'' is precisely analagous to the representation independence properties that underly abstraction theorems for module systems based on  abstract types and will similarly serve to ensure that the type invariants maintained by $\tcvar{rstr}$ are conserved when new tycons are defined. In this case the invariant is that only strings that are in the regular language specified in the type index can be generated from an external term of regular string type. Note that $\texttt{"TEST"}$ is not in the regular language specified by $\concrx{[A-Z]+ \digit\digit\digit\digit}$, so it is critical that it not be allowed, despite being consistent with the type translation of $\stx{conf}$, i.e. it is a string. Such an invariant could not be maintained if the type system were treated as merely a ``bag of rules''.\todo{put conservativity here?} %We consider this more rigorously in Sec. \ref{metatheory}.

\section{Related and Future Work}\label{prior-work}
As discussed throughout the paper, our treatment of concrete syntax defers to recent work on \emph{type-specific languages} \cite{TSLs}. %As in our work, TSLs determine the meaning of introductory terms based on the type the term is analyzed against, but TSLs only support syntactic elaboration to a fixed semantics.

%\paragraph{Term Rewriting}
Language-integrated static term rewriting  (``macro'') systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible, but cannot extend the type system directly.% If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g. Scala recommends using only its ``black box'' macros, which  enforce a function-like typing discipline). These do not permit extensions to the underlying type system.
%Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities).

@@

%\paragraph{Optimization}
When a simple embedding that preserves the static semantics exists, but a different  embedding is needed to preserve a desired cost semantics, term rewriting  can also be used to perform ``optimizations'', achieving an isomorphic embedding. Care is needed when this changes the type of a term to ensure that invariants are preserved. Type abstraction has been used for this purpose in work on \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}. This does not enable new type systems, i.e. in \cite{Rompf:2012:LMS}, Scala's type system must suffice. %As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order. %The metaprograms can modularly be shown not to violate the type system of the host language, Modular reasoning is possible as long as dispatch is explicit (e.g. macros) or based on an unambiguous language mechanism (e.g. trait composition in LMS) and made simpler by the fact that transformations must be type-preserving.

>>>>

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\chi}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}\vspace{-8px}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}. This form is analagous to the introductory form for kind $\kity$ described in Sec. \ref{sec:type-translations}, $\sqity{\qity}$. Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and dynamics simply recurse through these in the same manner as shown in Sec. \ref{sec:type-translations}. There is also an analagous unquote form, $\quq{\st}$. %The supplement gives the analagous rules.

The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$. These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.  Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$. We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
\[\small
\begin{array}{l}
\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}~\keyw{in}\\
\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
\quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx} ~\keyw{in}\\
\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}~\keyw{in}\\
\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}~\keyw{in}\\
\quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
}\end{array}
\]

@@

The \emph{argument interface list} is a static list of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$. It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$.

@@

\inferrule[k-syn]{
    n' < n
}{
    \sofkX{\ssyn{n'}}{\kty\times\kitm}
}
\end{mathpar}
The following lemma characterizes thus characterizes the argument list interface generation judgement:\vspace{-3px}
\begin{lemma}
\small If $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
\end{lemma}\vspace{-3px}
The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$. This is to prevent out-of-bounds errors: tycon providers do not write these forms directly, only accessing them via the argument interface list, which is guaranteed to have the correct length. This is  safe because $n$ is only an upper bound, so it can be relaxed safely.

@@

This judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement. The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous):
\begin{mathpar}\small
\inferrule[abs-anatrans-new]{
    n \notin \text{dom}(\memG)\\
    \keyw{nth}[n](\es) = e\\
    \eanaX{e}{\st}{\iota}\\
    \tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\ity}{\memD'}\\
    (x~\text{fresh})
}{
    \edeabs{c}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
}
\end{mathpar}

@@

Finally, the twelfth premise checks the abstracted term translation against the abstracted type translation. Here, we are checking $\Delta_\text{ex/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.: 
\[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\]
In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively. But because our  definition treated arguments parametrically, the check succeeds.% We  describe  an ill-behaved operator in Sec. \ref{metatheory}.

@@

The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstracted  term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.

@@

\begin{theorem}[Static Preservation]\label{thm:static-preservation}
If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
\end{theorem}
The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be mutually defined. %The mutual induction is well-founded because the total number of intro and targeted sub-terms being considered decreases.
\begin{theorem}[Type Synthesis]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
\end{theorem}
\paragraph{Type Safety}
Type safety in a typed translation semantics requires that well-typed external terms translate to well-typed internal terms. Type safety for the IL \cite{pfpl} then implies that evaluation cannot go wrong. To prove this, we must prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}:% produced as the translation of the external type. Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.

\begin{theorem}[Type-Preserving Translation]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
\end{theorem}
\begin{proof-sketch}
The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption. The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\emptyset \vdash \delta : \Delta_\text{abs}$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e. that all variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and so applying them gives a well-typed term.
\end{proof-sketch}

@@

\begin{tyconinvariant}[Regular String Soundness]
If $\eana{\emptyset}{\Phi_\text{rstr}}{e}{\sty{\tcvar{rstr}}{\concrx{r}}}{\iota}$ and $\iota \Downarrow \iota'$ then $\iota'=\texttt{"s"}$ and $\texttt{"s"}$ is in the regular language $\mathcal{L}(\texttt{r})$.
\end{tyconinvariant}
\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can essentially treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$. Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}. The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type. But if so, the abstracted translation must have type $\tau_\text{abs}=\alpha$. Thus, invariants cannot be violated by relational parametricity in the IL (i.e. a ``free theorem'') \cite{WadlerThms}. \end{proof-sketch}

The reason why (syn-targ-other) is never a problem in proving a tycon invariant, i.e. \emph{translation independence} of tycons, turns out to be the same reason extending the tycon context conserves all tycon invariants. A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g. $\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g. $\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, which would fail. %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
 We can state this more generally:

\begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
\item If $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\sttyidx, \iota)$.
\end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\sttyidx, \iota)$.
\end{itemize}
(if proposition $P(\sttyidx, \iota)$ is \emph{modular}, defined below)
\end{theorem}
\begin{proof-sketch}
The proof maps every well-typed term under $\Phi'$ to a well-typed term under $\Phi$ with the same translation, and if the term has a type constructed by a tycon in $\Phi$, e.g. $\tc$, the mapping still has a type constructed by that tycon with the same type translation. If $P(\sttyidx, \iota)$ is preserved under the mapping over $\sttyidx$ then we can simply invoke the assumed tycon invariant. We call such propositions \emph{modular}. Non-modular propositions are uninteresting because the mapping's effect on static terms is only to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for $m$ uniquely associated with $\tc'$. Any propositions decidable by the SL are modular. On external terms, the mapping replaces all intro and targeted terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices. Typing, kinding, static normalization and selective translation abstraction are preserved under the mapping, defined inductively in the supplement. \end{proof-sketch}
%Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$. Had we not gone through the machinations of holding types and terms abstract, however, the check would succeed, because $\keyw{str}$ would no longer have been held abstract as $\alpha_1$. This form of ``type translation independence'' is precisely analagous to the representation independence properties that underly abstraction theorems for module systems based on  abstract types and will similarly serve to ensure that the type invariants maintained by $\tcvar{rstr}$ are conserved when new tycons are defined. In this case the invariant is that only strings that are in the regular language specified in the type index can be generated from an external term of regular string type. Note that $\texttt{"TEST"}$ is not in the regular language specified by $\concrx{[A-Z]+ \digit\digit\digit\digit}$, so it is critical that it not be allowed, despite being consistent with the type translation of $\stx{conf}$, i.e. it is a string. Such an invariant could not be maintained if the type system were treated as merely a ``bag of rules''.\todo{put conservativity here?} %We consider this more rigorously in Sec. \ref{metatheory}.

\section{Related Work and Discussion}\label{prior-work}
%\paragraph{Term Rewriting}
Language-integrated static term rewriting systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible. % If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g. Scala recommends using only its ``black box'' macros, which  enforce a function-like typing discipline). These do not permit extensions to the underlying type system.
%Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities). 
Similarly, when an embedding that preserves a desired static semantics exists, but a different  embedding preserves the cost semantics, term rewriting  can also be used to perform ``optimizations'', achieving an isomorphism. Care is needed when this changes the type of a term. Type abstraction has been used for  this purpose in \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}. In both cases, the type system is fixed (e.g. in LMS, Scala's).

@@

%\paragraph{Optimization}
 %This does not enable new type systems, i.e. in \cite{Rompf:2012:LMS}, Scala's type system is used. %As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order. %The metaprograms can modularly be shown not to violate the type system of the host language, Modular reasoning is possible as long as dispatch is explicit (e.g. macros) or based on an unambiguous language mechanism (e.g. trait composition in LMS) and made simpler by the fact that transformations must be type-preserving.
